---
title: "Mortality predictors in hematological patients, admitted to intensive care unit: a single-centre observational study"
author: "Sergei Vladimirov"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	error = FALSE,
	message = FALSE,
	warning = FALSE
)

library(blorr)
library(stargazer)

library(tidyverse)
library(magrittr)
library(hms)
library(RSQLite)
library(DBI)
library(kableExtra)
library(ggsci)
library(patchwork)
library(tufte)
library(ggthemes)
library(ggplot2)
library(finalfit)
library(lubridate)
library(janitor)
library(readxl)
library(readr)
library(hrbrthemes)
library(flextable)
library(gtsummary)
#install.packages('pROC')
library(pROC)
library(GGally)
#install.packages('DiagrammeR')
library(DiagrammeR)
#install.packages('sjPlot')
library(sjPlot)


```

# Abstract

Purpose:
Previous research showed conflicting results about mortality predictors in critically ill patients with hematological malignancies (HM). The primary aim of this study was to determine mortality in critically ill patients with HM in intensive care unit (ICU) and to reveal risk factors predicting the outcome.

Methods and materials:
All patients with HM admitted to ICU at our hospital during 1 year were enrolled. Clinical data upon ICU admission was collected and then outcomes were estimated.

Results:
ICU mortality was 52% among 98 HM patients.
According to multivariate analysis, male sex (OR 2.75, CI: 1.15–6.88, p = 0.026), high SOFA score (OR 1.30, CI: 1.07–1.62, p = 0.012) and high Charlson Comorbidity Index (OR 1.30, CI: 1.06–1.64, p = 0.016) were associated with the outcome.

Conclusion:
We found three independent predictors for ICU mortality. Further research is
needed to validate current findings and reveal new mortality predictors in critically ill HM patients.

```{r echo=FALSE, include = FALSE }
#  Importing and fixing final dataset

original_data <- read_csv('final_project_df.csv')


original_data <- original_data %>%
  mutate_if(is.character, as.factor) %>%
  mutate(neutropenia = as.factor(neutropenia)) %>% 
   mutate(
    resp_insuf = as.factor(case_when(
      sofa_resp > 0 ~ 1,
      sofa_resp < 1 ~ 0
    )),
    neur_def = as.factor(case_when(
      sofa_gcs > 0 ~ 1,
      sofa_gcs < 1 ~ 0
    )),
    hep_insuf = as.factor(case_when(
      sofa_bil > 0 ~ 1,
      sofa_bil < 1 ~ 0
    )),
    hemod_instab = as.factor(case_when(
      sofa_map > 0 ~ 1,
      sofa_map < 1 ~ 0
    )),
    thrombocytopenia = as.factor(case_when(
      sofa_plt > 0 ~ 1,
      sofa_plt < 1 ~ 0
    )),
    renal_insuf = as.factor(case_when(
      sofa_renal > 0 ~ 1,
      sofa_renal < 1 ~ 0
    ))
    ) %>% 
  select(id, gender, age, cci, diagnosis,
         disease_status, days_bfr_icu, 
         sofa, neutropenia, is_deceased)


```


# Introduction

Patients with hematologic malignancies (HMs) have an increased risk of death compared to other oncology patients in settings of intensive care unit (ICU). According to previous research1, the predictive value of different admission factors is constantly changing, and several studies on this topic have conflicting results.
Our research question is addressed to identifying factors in the condition of hematological patients that can predict a poor ICU outcome. The primary aim of this study was to determine ICU mortality and risk factors predicting the outcome of critically ill patients with HMs.
We hypothesized that there are some significant factors in the condition of HM patients at the moment of ICU admission that are associated with mortality.

# Methods

Our research is an observational, single-center study.
We retrospectively evaluated baseline characteristics of adult HM patients non-electively admitted to the medical ICU of our hospital from July 1, 2022 to June 30,2023. All the HM diagnoses were based on the pathological examination results or medical records. All data were obtained from a local ICU electronic database.
Cases with a critical amount of missing data, as well as those who died or
transferred to the floor within 24 hours were excluded from the study.

```{r}
df_variable <- colnames(original_data)

var_definition <- c(
  "anonymized identical number of patient",
  "gender of patient",
  "age of patient",
  "Charlson Comorbidity Index - most widely used scoring system for comorbidities",
  "type of hematological malignancy",
  "status of HM progression",
  "number of days in hospital before admission to ICU",
  "sequential organ failure assessment score (SOFA score), is used  to determine the extent of organ function or rate of failure",
  "absolute neutrophil count < 1000 per mm3",
  "outcome (1 is death, 0 is survival)")

var_type <- c(
  "numerical",
  "categorical, binary",
  "numerical, years",
  "numerical, points",
  "categorical",
  "categorical: p (progression) - relapsed or refractory disease, n - newly diagnosed",
  "numerical, days",
  "numerical, points",
  "categorical, binary (1 = yes, 0 = no)",
  "categorical, binary (1 is death, 0 is survival)")



var_status <-  c(
  "independent",
  "independent",
  "independent",
  "independent",
  "independent",
  "independent",
  "independent",
  "independent",
  "independent",
  "dependent")
  
var_df <-  
bind_cols(df_variable, var_definition, var_type, var_status) %>% 
  rename( variable = '...1', definition = '...2', type = '...3', status = '...4')



var_df %>% kable()
#%>% flextable()
#  theme_box() %>% 
#  set_caption('A list of variables used in analysis')

```


Upon ICU admission, we included baseline patients’ information on demographic data, diagnosis, extent of organ dysfunction, comorbidity burden, presence of neutropenia, disease status and number of days in hospital prior to ICU referral. Based on above-mentioned information, independent variables were created, and the outcome of ICU hospitalization was set as a dependent variable (Table 1).
This study was carried out according to the principles of the Declaration of Helsinki. Approval was granted by the local ethics committee (date: 03.08.2023 / no: 1236). Since we performed a retrospective analysis of routinely collected de-identified data, informed consents from the patients were not required. The trial was registered.
All statistical analyses were performed using R version 4.2.1 (R Core Team, 2022).
Сontinuous variables were described as mean (standard deviation), median (25–75 percentiles) and range. We performed a univariate analysis first to calculate the odds ratio (OR) of mortality, and statistically significant factors were then used in a multivariate logistic regression model to determine outcome prediction. Receiver operating curve (ROC) was done for the final model. In this research, all the tests were two-sided, and p < 0.05 was considered as statistically significant.


```{r echo=FALSE, include= TRUE, }
#  Importing and fixing final dataset

original_data <- read_csv('final_project_df.csv')


original_data <- original_data %>%
  mutate_if(is.character, as.factor) %>%
  mutate(neutropenia = as.factor(neutropenia)) %>% 
   mutate(
    resp_insuf = as.factor(case_when(
      sofa_resp > 0 ~ 1,
      sofa_resp < 1 ~ 0
    )),
    neur_def = as.factor(case_when(
      sofa_gcs > 0 ~ 1,
      sofa_gcs < 1 ~ 0
    )),
    hep_insuf = as.factor(case_when(
      sofa_bil > 0 ~ 1,
      sofa_bil < 1 ~ 0
    )),
    hemod_instab = as.factor(case_when(
      sofa_map > 0 ~ 1,
      sofa_map < 1 ~ 0
    )),
    thrombocytopenia = as.factor(case_when(
      sofa_plt > 0 ~ 1,
      sofa_plt < 1 ~ 0
    )),
    renal_insuf = as.factor(case_when(
      sofa_renal > 0 ~ 1,
      sofa_renal < 1 ~ 0
    ))
    ) %>% 
  select(id, gender, age, cci, diagnosis,
         disease_status, days_bfr_icu, 
         sofa, neutropenia, is_deceased)



```



