---
title: "Mortality predictors in heamtological patients, admitted to intensive care unit: a single-centre observational study"
author: "Sergei Vladimirov"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	error = FALSE,
	message = FALSE,
	warning = FALSE
)

#install.packages('hrbrthemes')

library(tidyverse)
library(magrittr)
library(hms)
library(RSQLite)
library(DBI)
library(kableExtra)
library(ggsci)
library(patchwork)
library(tufte)
library(ggthemes)
library(ggplot2)
library(finalfit)
library(lubridate)
library(janitor)
library(readxl)
library(readr)


pacman::p_load(
  rio,          # File import
  here,         # File locator
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  gtsummary,    # summary statistics and tests
  rstatix,      # summary statistics and statistical tests
  janitor,      # adding totals and percents to tables
  scales,       # easily convert proportions to percents  
  flextable,
  epicalc
  )


```

# Abstract

Despite progress in therapeutic options and improved prognosis of hematological
patients in recent years, the mortality of patients with hematological malignancies
(HM) requiring admission to intensive care units (ICU) remains high. According to
the results of the latest observational research, the factors of poor outcome in such
patients vary.
I am going to identify prognostic factors associated with the ICU and hospital
mortality in the cohort of HM patients who were admitted to ICU in my hospital
between April 2022 and April 2023.
My dataset contains 43 variables of 69 consecutive critically ill patients. The
following parameters were recorded within 24 h of ICU admission: demographic
data, diagnosis, some major laboratory data, values of severity scales (SOFA and
SAPS), indicators of vital functions. Three columns contain information about
outcomes (dead_icu - outcome of ICU case, dead_hospital - outcome of overall
hospitalization, icu_los - length of stay in ICU).
I am planning to perform a regression analysis in order to reveal some meaningful
factors related to ICU and hospital mortality in these patients.
Source of data: electronic health records.

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
Reference Style Links and Images

```{r echo=FALSE, include= TRUE, }
#  Importing and fixing final dataset

original_data <- read_csv('final_project_df.csv')

original_data <- original_data %>% 
  mutate_if(is.character, as.factor) %>%
  mutate(neutropenia = as.factor(neutropenia)) %>% 
  select(!c(age_group, age_group_20, septic_shock, on_mv))

head(original_data)



```



